---
title: "Combined Nephron Trajectory Analysis"
output:
    workflowr::wflow_html:
        code_folding: hide
        df_print: paged
---

```{r knitr, include = FALSE}
DOCNAME = "07C_Combined_Trajectory"
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = TRUE,
                      cache.path     = paste0("cache/", DOCNAME, "/"),
                      cache.comments = FALSE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.width      = 10,
                      fig.height     = 8,
                      message        = FALSE,
                      warning        = FALSE)
```

```{r libaries, cache = FALSE}
# scRNA-seq
library("Seurat")
library("monocle")

# Plotting
library("viridis")

# Presentation
library("glue")
library("knitr")

# Parallel

# Paths
library("here")

# Output
library("jsonlite")

# Tidyverse
library("tidyverse")
```

```{r source, cache = FALSE}
source(here("R/output.R"))
```

```{r paths}
comb.neph.path <- here("data/processed/Combined_nephron.Rds")
```

Introduction
============

In this document we are going to take the results of reclustering the nephron
lineage in the combined analysis and perform pseudotime trajectory analysis
using Monocle. This should tell us more about how those clusters relate to each
other and what is happening during the developmental process.

```{r load, cache.extra = tools::md5sum(comb.neph.path)}
if (file.exists(comb.neph.path)) {
    comb.neph <- read_rds(comb.neph.path)
} else {
    stop("Clustered Combined nephron dataset is missing. ",
         "Please run '07C_Combined_Nephron.Rmd' first.",
         call. = FALSE)
}

comb.neph.cds <- importCDS(comb.neph)
comb.neph.cds <- estimateSizeFactors(comb.neph.cds)
```

Gene selection
==============

We first need to select a set of genes to use for this analyis. As we have
already identified marker genes for each cluster it makes sense to use those.
We will use the top 100 conserved marker genes for each cluster with the
greatest absolute foldchanges. 

```{r select-genes}
con.markers <- read_csv(here("output/07B_Combined_Nephron",
                             "conserved_markers.csv"))

ordering.genes <- con.markers %>%
    group_by(cluster) %>%
    arrange(-abs(mean_avg_logFC)) %>%
    top_n(100, mean_avg_logFC) %>%
    pull(gene) %>%
    unique()

comb.neph.cds <- setOrderingFilter(comb.neph.cds, ordering.genes)
fData(comb.neph.cds)$ExprsCells <- rowSums(as.matrix(exprs(comb.neph.cds)) > 0)
```

Taking the intersection of these sets gives us `r length(ordering.genes)`
selected genes.

Order cells
===========

Monocle then uses the selected genes to perform dimensionality reduction using
the DDRTree algorithm and then orders the cells along a trajectory.

```{r order-cells}
comb.neph.cds <- reduceDimension(comb.neph.cds, max_components = 2,
                                 method = "DDRTree")
comb.neph.cds <- orderCells(comb.neph.cds)

root.state <- 4
n.branch <- length(
    comb.neph.cds@auxOrderingData[[comb.neph.cds@dim_reduce_type]]$branch_points
)
n.states <- length(unique(pData(comb.neph.cds)$State))

comb.neph.cds <- orderCells(comb.neph.cds, root_state = root.state)
```

Plot trajectory {.tabset}
---------------

```{r plot-trajectory, results = "hide"}
color.bys <- c("State", "Pseudotime", "NephCluster", "Dataset")

src_list <- lapply(color.bys, function(color.by) {
    src <- c("### {{color.by}} {.unnumbered}",
             "```{r plot-trajectory-{{color.by}}}",
             "plot_cell_trajectory(comb.neph.cds, color_by = '{{color.by}}')",  
             "```",
             "")
    knit_expand(text = src)
})

out <- knit_child(text = unlist(src_list), options = list(cache = FALSE))
```

`r out`

Crossover
---------

Monocle assigns each cell a "state" which corresponds to a segment of the tree.
In this case Monocle has identified `r n.branch` branch points and `r n.states` 
states. We have chosen state `r root.state` to be the root state of our tree.
Let's look at how those states relate to our nephron clusters.

```{r crossover}
pData(comb.neph.cds) %>%
    count(State, NephCluster) %>%
    complete(State, NephCluster, fill = list(n = 0)) %>%
    group_by(State) %>%
    mutate(Total = sum(n)) %>%
    ggplot(aes(x = NephCluster, y = n / Total, fill = State)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    facet_wrap(~ State, ncol = 1, labeller = label_both) +
    labs(title = "State by nephron cluster",
         x = "Nephron cluster",
         y = "Percentage of cells in state") +
    theme_minimal()
```

Pseudotime genes
================

Heatmap
-------

We can show the expression of genes along this pseudotime trajectory. Here is a
heatmap showing a selection of interesting kidney organoid genes.

```{r select-plot-genes}
plot.genes <- c("PECAM1", "CDH5", "MEIS1", "PDGFRA", "HMGB2", "CENPA", "SIX1",
                "DAPL1", "NPHS1", "PODXL", "S100A8", "TYROBP", "MAL", "EMX2",
                "LRP2", "GATA3", "SLC12A1", "SPINT2", "TUBB2B", "STMN2",
                "TTHY1", "HBA1", "HBG1")

is_present <- plot.genes %in% rownames(comb.neph.cds)
is_exprs <- fData(comb.neph.cds)[plot.genes[is_present], "ExprsCells"] > 10
plot.genes.use <- plot.genes[is_present][is_exprs]
```

> The following genes aren't present in this dataset and will be skipped:
> `r glue_collapse(plot.genes[!is_present], sep = ", ", last = " and ")`

> The following genes aren't expressed in at least 10 cells in this dataset and
> will be skipped:
> `r glue_collapse(plot.genes[!is_present][!is_exprs], sep = ", ", last = " and ")`

```{r time-genes-heatmap}
plot_pseudotime_heatmap(comb.neph.cds[plot.genes.use, ],
                        num_clusters = 3,
                        cores = 1,
                        show_rownames = TRUE,
                        hmcols = viridis(100))
```

Expression {.tabset}
----------

We can also show the changes in expression of each gene individually.

```{r time-genes-exprs, results = "hide"}
src_list <- lapply(color.bys, function(color.by) {
    src <- c("### {{color.by}} {.unnumbered}",
             "```{r time-genes-exprs-{{color.by}}}",
             "plot_genes_in_pseudotime(",
                 "comb.neph.cds[plot.genes.use, ],",
                 "ncol = 2, color_by = '{{color.by}}')",  
             "```",
             "")
    knit_expand(text = src)
})

out <- knit_child(text = unlist(src_list), options = list(cache = FALSE))
```

`r out`

Branching genes
===============

Monocle can identify branch points and it can be useful to examine how genes are
expressed differentially along each branch.

Heatmap {.tabset}
-------

In this heatmap the centre shows the branch point, one branch moves towards the
left and the other towards the right.

### Branch point 1 {.unnumbered}

```{r branch-genes-heatmap-1, results = "hide"}
plot_genes_branched_heatmap(comb.neph.cds[plot.genes.use, ],
                            branch_point = 1,
                            trend_formula = '~ sm.ns(Pseudotime, df = 4) * Branch',
                            num_clusters = 3,
                            show_rownames = TRUE,
                            hmcols = viridis(100))
```

### Branch point 2 {.unnumbered}

```{r branch-genes-heatmap-2, results = "hide"}
plot_genes_branched_heatmap(comb.neph.cds[plot.genes.use, ],
                            branch_point = 2,
                            num_clusters = 3,
                            show_rownames = TRUE,
                            hmcols = viridis(100))
```

Expression {.tabset}
----------

We can also look at how individual genes change from a branch point. Here the
left is the branch point and the right is the end of each branch. Expression
along a branch is shown by the two different lines.

### Branch point 1 {.unnumbered .tabset}

```{r branch-genes-exprs-1, results = "hide"}
src_list <- lapply(color.bys, function(color.by) {
    src <- c("#### {{color.by}} {.unnumbered}",
             "```{r branch-genes-exprs-1-{{color.by}}}",
             "plot_genes_branched_pseudotime(comb.neph.cds[plot.genes.use, ],",
                 "branch_point = 1,",
                 "trend_formula = '~ sm.ns(Pseudotime, df = 4) * Branch',",
                 "ncol = 2, color_by = '{{color.by}}')",  
             "```",
             "")
    knit_expand(text = src)
})

out <- knit_child(text = unlist(src_list), options = list(cache = FALSE))
```

`r out`

### Branch point 2 {.unnumbered .tabset}

```{r branch-genes-exprs-2, results = "hide"}
src_list <- lapply(color.bys, function(color.by) {
    src <- c("#### {{color.by}} {.unnumbered}",
             "```{r branch-genes-exprs-2-{{color.by}}}",
             "plot_genes_branched_pseudotime(comb.neph.cds[plot.genes.use, ],",
                 "branch_point = 2,",
                 "ncol = 2, color_by = '{{color.by}}')",  
             "```",
             "")
    knit_expand(text = src)
})

out <- knit_child(text = unlist(src_list), options = list(cache = FALSE))
```

`r out`

Summary
=======

Parameters
----------

This table describes parameters used and set in this document.

```{r parameters}
params <- toJSON(list(
    list(
        Parameter = "n.genes",
        Value = length(ordering.genes),
        Description = "Number of genes used for trajectory analysis"
    ),
    list(
        Parameter = "n.states",
        Value = n.states,
        Description = "Number of states identified by trajectory analysis"
    ),
    list(
        Parameter = "n.branch",
        Value = n.branch,
        Description = "Number of branch points identified by trajectory analysis"
    ),
    list(
        Parameter = "root.state",
        Value = root.state,
        Description = "State chosen as root of the tree"
    )
), pretty = TRUE)

kable(fromJSON(params))
```

Output files
------------

This table describes the output files produced by this document. Right click
and _Save Link As..._ to download the results.

```{r write-data}
write_rds(comb.neph.cds, here("data/processed/Combined_trajectory.Rds"))
```

```{r output}
dir.create(here("output", DOCNAME), showWarnings = FALSE)

write_lines(params, here("output", DOCNAME, "parameters.json"))

kable(data.frame(
    File = c(
        glue("[parameters.json]({getDownloadURL('parameters.json', DOCNAME)})")
    ),
    Description = c(
        "Parameters set and used in this analysis"
    )
))
```
